It's all about Meaning!\\{\normalsize A Theory for Data Migration during Software Development

### 2022-12-11 2022 Migration/articleMigration.tex
Repair the first example for the new requirements.

\section{Introduction}
\label{sct:Introduction}
   To automate the datamigrations that come with incrementally developing and deploying information systems.
   That is the purpose of the theory proposed in this paper.

   We believe in generating information systems%
\footnote{In the sequel, the word ``system'' refers to the phrase ``information system''. This simplifies the language a little. }
   to prevent human programming errors.
   For, a mistake not made can never propagate into production.
   We also believe that information systems should be released into production in small increments,
   to obtain shorter release cycles, more frequent releases, and thus a more agile software development process%
   \footnote{These effects can be measured in terms of DevOps metrics such as
   deployment frequency,
   reliability of deployments, and
   change failure rate~\cite{DevOps2021}.}.
   By installing upgrades automatically, frequently, in small increments, and without down-time,
   the users experience a system that evolves organically, without even noticing the updates.
   Thus, we believe our theory will help to automate data migrations correctly and reliably.
   The benefits work in two directions.
   Automation data migrations saves effort and errors,
   contributing to more frequent and smaller releases.
   Smaller and faster releases also mean smaller migrations
   and therefore a smaller difference between the old system and the new system.
   The smaller that difference, the simpler the migration and the more of it can be automated.

   In each increment, we distinguish an ``old system'' and a ``new system''.
   The difference between the old system and the new system is not just a difference in two schemas.
   When deploying the increment, the data from the old system needs to migrate to the new system as well.
   Incremental evolution poses specific requirements to data migration,
   which sets it apart from data migrations for other purposes.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers may exclude new functionality to avoid introducing new errors in an otherwise error-prone migration process.
   In our work, however, introducing new functionality and removing obsolete functionality is the very purpose of an increment.
   Data migration literature sometimes uses a different notion of increment.
   For example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014} define an increment as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   In our work, however, we want to get rid of the old system and not retain it.
   So, we need to look at a different approach.
   Examples like these have convinced us to define a new method for data migration that is specifically meant for systems in production that are
   continually evolving in small increments.

@@

\section{Analysis}
   Information systems are typically used by distributed actors (both users and computers) who continually work with data.
   Every system contains a dataset, which represents the state of the system.
   Events that the system detects may cause the state to change.
   To keep our theory technology independent, we will assume that a dataset contains triples.
   This makes our theory valid for any kind of database that triples can represent,
   such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.

@@

   We will assume that the schema of a system contains constraints on its dataset.
   These constraints represent the semantics of the system, similar to the theory developed by Spivak~\cite{Spivak2012}.
   This allows us to be explicit about ``preserving the meaning as much as possible''.
   In contrast with Spivak, however, our approach does not work with functions (functors)
   but with relations similar to Alloy~\cite{Alloy2006} and allegories%
\footnote{Allegories are a specific type of categories, which the reader does not need to understand for the purpose of this paper}~\cite{Zielinski2013}.

   In this work we use the Ampersand compiler~\cite{Joosten-JLAMP2018},
   to generate information systems.
   A script in the Ampersand language represents the system's semantics by constraints only.
   We don't have to formalize the semantics from specifications or from source code
   because the constraints are already explicitly available in the Ampersand script.
   A system generated by the Ampersand compiler will keep these constraints satisfied at all times.
   This makes Ampersand an obvious platform for validating our theory.
   An Ampersand script contains just enough information to generate a complete system,
   which means that a classical database schema (i.e. data structure plus semantics) can be extracted from the Ampersand script.
   For the purpose of this paper we can equate an Ampersand script with the schema of the generated system.

   For data migration, complications arise from accumulated data pollution in the old system.
   We distinguish data pollution in the following categories:
\begin{itemize}
   \item data pollution that cannot be captured in constraints on the dataset,
   such as a street address that has become obsolete because the system never registered the change of address.
   \item data pollution that violates a constraint in the schema.
   \item data pollution that violates a constraint that is not in the schema.
   \item data pollution that is a consequence of an erroneous constraint in the schema.
\end{itemize}
   The first category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
   In some cases, constraints on data can be formulated,
   so the system can help the user and simplify (or eliminate) the corresponding working procedure.
   So, such constraints may help to move this type of pollution to one of the other categories.
   The second category is about violations of constraints from the schema.
   Such violations do not occur in our theory because the old system was generated by Ampersand,
   so that system will keep all constraints from the schema satisfied at all times.
   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
   The third category, data pollution that is not captured by any of the constraints mentioned in the schema,
   can be dealt with by adding constraints in the new system.
   In that case, a migration engineer must ensure that the new system can start with data that satisfies all constraints in its schema.
   So some human intervention may be necessary.
   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in the new system.

\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.35]{Migration.png}
   \end{center}
\caption{Data migration}
\label{fig:event flow}
\end{figure}
   This illustrates that automating data migrations has its limits.
   Some effort of a migration engineer can never be excluded.
   If the semantics encoded in the new schema differs from the old schema,
   a migration engineer must make choices.
   If the interpretation of data changes,
   a migration engineer must remove obsolete constraints from the old system and introduce new constraints in the new system.
   The consequences of such changes may require changes in data as well.
   The migration engineer may even want to adapt existing data of which the constraints in the schema have not changed.
   All of this requires a third schema, which specifies the migration itself.
   This migration schema preserves data from the old system, it may introduce new data automatically (generated from the old dataset),
   and it may require users to introduce new data (which will take time). 
   It must contain the old schema, but it also contains the new schema and possibly some new data, which is specified by the migration engineer.
   As the migration schema must distinguish between the old dataset and the new dataset,
   the migration must be based on the disjoint union of the two.
   So the idea of a migration schema is to compute the disjoint union of the old and the new schemas
   and to present it to the migration engineer in the form of source code,
   so the migration engineer can change everything she wants to suit the particulars of the data migration.

@@

   Atoms serve as data elements.
   They are values without internal structure of interest, meant to represent data elements in a database.
   From a business perspective, atoms represent concrete items of the world,
   such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
   By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
   All atoms are taken from an infinite set called $\Atoms$.

@@

   The relation $\inst:\Pair{\Atoms}{\Concepts}$ is the instance relation between atoms and concepts.
   The expression $a\ \inst\ C$ means that atom $a$ is an \emph{instance} of concept $C$.
   This relation is used in the type system.
   In the type system, every atom must have a type,
   which makes $\inst$ a total relation:
\begin{eqnarray}
   \forall a\in\atoms\ \exists C\in\Concepts: a\ \inst\ C
\label{eqn:concepts}
\end{eqnarray}

@@

   Triples serve to represent data.
   A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
   A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
\begin{eqnarray}
   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
   \label{eqn:type}
\end{eqnarray}
   For example, $\triple{\text{\atom{Peter}}}{\declare{\id{phone}}{\tt Person}{\tt TelephoneNumber}}{\text{\atom{074238991}}}$ is a triple.
   Equation~\ref{eqn:type} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
   In practice, users can say that Peter has telephone number 074238991.
   So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
   This ``meaning from practice'' has no consequences in the formal world.
   We leave it entirely up to a user to attach a practical meaning to a triple.
   In this paper, the informal meaning is interesting only for the sake of illustration.

   $\triples$ is a finite set of triples and $\inst$ is the instance relation.
   In case we need to be explicit about the dataset,
   we will write $\triples_{\dataset}$ and $\inst_{\dataset}$ to refer to $\triples$ and $\inst$ respectively.
   Every dataset is an element of an infinite set called $\Dataset$.
   To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.
   Every atom has the type of the relation in which it resides:

@@

   In order to do static type checking,
   all concepts and relations must be ``known'' in the schema:
\begin{eqnarray}
   \declare{n}{A}{B}\in\rels&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
   \label{eqn:relationsIntroduceConcepts}\\
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels\ \wedge\ a\ \inst\ A\wedge b\ \inst\ B
   \label{eqn:define R}
\end{eqnarray}

   A rule in a schema serves to constrain the dataset,
   to ensure the semantic integrity of the dataset.
   For every rule $u$ in the schema, we assume there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
   that represents the set of violations of rule $u$ in any dataset.
   We also assume a function $\Sign{u} : \Dataset\rightarrow\Pair{\Concepts}{\Concepts}$
   that represents the signature of a rule, i.e. the type of atoms in violations:
\begin{eqnarray}
   \sign{u}{\dataset}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
   \label{eqn:typed violations}
\end{eqnarray}
   To determine whether rule $u$ is satisfied in dataset $\dataset$,
   we define $\sat{u}{\Dataset}$:
\begin{eqnarray}
   \sat{u}{\dataset}&\Leftrightarrow&\viol{u}{\dataset}=\emptyset
   \label{eqn:sat}
\end{eqnarray}
   Violations are used to produce meaningful error messages and to trigger changes in the dataset to restore satisfaction of that rule.

   We assume a special kind of rule that is used for specialization.
   The rule $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
\begin{equation}
   \label{eqn:specialization}
   \sat{A\isa B}{\dataset}\ \Leftrightarrow\ \forall a: a\ \inst\ A\rightarrow a\ \inst\ B
\end{equation}
   We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
   All specialization rules in a schema together form a partial order $\isa : \Pair{\concepts}{\concepts}$.
   Specialization rules are special for two reasons:
   Firstly, the system keeps $A\isa B$ satisfied automatically by changing $\inst$ whenever needed.
   This means that a system that contains rule $A\isa B$ can never be brought into a state in which $\id{sat}_{A\isa B}$ is False.
   This allows for certain run-time optimizations, which are beyond the scope of this paper.
   A second property that sets specialization rules apart, is that they play a role in the Ampersand compiler.
   For example, code that determines equality between atoms only works for atoms that are an instance of the same concept.
   To ensure that this does not result in problems at run-time, users are required to write all specialization rules explicitly.

@@

\subsection{Information Systems}
\label{sct:Information Systems}
   The purpose of an information system is to make data meaningful to its users.
   Users have their own tasks and responsibilities
   and may work from different locations and on different moments.
   This collective use serves a purpose which we loosely call ``the business''.
   As a consequence, the data in a system changes continually.
   To preserve meaning, users are trying to maintain semantic constraints on the data,
   amidst of all changes that are going on around them.

   This section introduces the notion of information system.
   As before, we will suffix the elements of this definition if the situation requires it.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\functionality,\roles,\maintain\ra$, in which
\begin{itemize}
   \item dataset $\dataset$ is defined as in section~\ref{sct:Datasets};
   \item schema $\schema$ is defined as in section~\ref{sct:Schemas};
   \item $\functionality$ represents the functionality of the system, which is typically software;
   \item $\roles$ is a set of roles;
   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$.
\end{itemize}
\end{definition}
   A \define{role} is a name that identifies a group of users.
   It serves as a placeholder for a person or a machine (i.e. an actor) who can change the dataset.
   The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.
   In the sequel, when we talk about a role we actually mean an actor who fulfills that role in the system.

   Every rule in a schema is meant to be kept satisfied invariably.
   The only way to break this invariance is to remove the rule from the schema.
   So the "normal" state of a system is that its dataset $\dataset$ contains no violations of any rule in the schema:
   $\forall u\in\rules:\ \sat{u}{\dataset}$.
   However, the system continually registers events that insert or delete triples in its dataset,
   which may cause rules to be satisfied no longer.
   This is a temporary breach of the invariance of the affected rules,
   which requires a reaction to restore invariance.
   Example: if a rule says that every request must be acknowledged, the receipt of a request will violate this rule.
   Sending a message that acknowledges this receipt will cause the rule to be satisfied again.

   In this paper we assume two ways of enforcing satisfaction of rules: automatic or manual.
   Manual enforcement is specified by the developer by assigning a rule $u$ to a role $o$.
   Any actor with role $o$, whether person or machine,
   can restore invariance of rule $u$ by inserting or deleting the right triples to compensate for the violations.
   For this reason, we assume that every rule is assigned to at least one role by means of the relation $\maintain$.
\begin{equation}
   \forall u\in\rules\ \exists o\in\roles:\ o\ \maintain\ u\label{eqn:maintain}
\end{equation}
   $o\ \maintain\ u$ means that role $o$ must keep rule $u$ satisfied, i.e.\ to keep $\sat{u}{\dataset}$ true.
   Not all ways of restoring invariance can be anticipated by an algorithm, which makes manual enforcement a necessary feature.
   As long as rule $u$ is not satisfied, its violations won't go away, stimulating the actor to do something about it.

   Automatic enforcement is specified by the developer with a special syntax,
   which describes not only the rule, but also the way to restore invariance.
   Whenever the dataset violates an enforce rule,
   the system will make other changes to the dataset to satisfy that rule.
   Hence, the violation of the rule exists during a small, finite amount of time in which the system restores its invariance.

   The fact that every rule has at least one role in the relation $\maintain$, means that all roles together are keeping all rules satisfied.

@@

   The example system also has a second relation that states which modules are part of which course.
   It is constrained by \verb-[UNI]- to be univalent.
   This means that every \verb-Module- is part of at most one \verb-Course-.
\begin{verbatim}
RELATION isPartOf[Module*Course] [UNI] =
[ ("Finance", "Management")
; ("Business Rules", "Business IT")
; ("Business Analytics", "Business IT")
; ("IT-Governance", "Management")
]
\end{verbatim}
   The third and last relation states which students are enrolled for which module.
   We leave it empty for now.
\begin{verbatim}
RELATION isEnrolledFor[Student*Module]
\end{verbatim}
   We also define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that student takes.
   In Ampersand, which is a syntactically sugared form of relation algebra,
   we give each rule a name and declare a role to preserve its invariance:
\begin{verbatim}
RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
ROLE Administrator MAINTAINS EnrollRule
\end{verbatim}
   The actual constraint, \verb#isEnrolledFor |- takes;isPartOf~#, is written in logic as:
\[\begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
\end{array}\]

>>>>

\section{Introduction}
\label{sct:Introduction}
   To automate the datamigrations that come with incrementally developing and deploying information systems;
   that is the purpose of the theory proposed in this paper.

   We believe in generating information systems%
\footnote{In the sequel, the word ``system'' refers to the phrase ``information system''. This simplifies the language a little. }
   to prevent human programming errors.
   For, a mistake not made can never propagate into production.
   We also believe that information systems should be released into production in small increments,
   to obtain shorter release cycles, more frequent releases, and thus a more agile software development process%
   \footnote{These effects can be measured in terms of DevOps metrics such as
   deployment frequency,
   reliability of deployments, and
   change failure rate~\cite{DevOps2021}.}.
   By installing upgrades automatically, frequently, in small increments, and without down-time,
   the users experience a system that evolves organically, without even noticing the updates.
   Thus, we believe our theory will help to automate data migrations correctly and reliably.
   The benefits work in two directions.
   Automation data migrations saves effort and errors,
   contributing to more frequent and smaller releases.
   Smaller and faster releases also mean a smaller difference between the old system and the new system.
   The smaller that difference, the simpler the migration can be and the more of it can be automated.

   In each increment, we distinguish an ``old system'' and a ``new system''.
   A difference of schemas complicates the migration of data between the old system and the new system.
   The purpose of incremental evolution poses specific requirements to data migration,
   which sets it apart from data migrations for other purposes.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers may exclude new functionality to avoid introducing new errors in an otherwise error-prone migration process.
   In our work, however, introducing new functionality and removing obsolete functionality is the very purpose of an increment.
   Data migration literature sometimes uses a different notion of increment.
   For example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014} define an increment as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   In our work, however, we want to get rid of the old system and not retain it.
   So, we need to look at a different approach.
   Examples like these have convinced us to define a new method for data migration that is specifically meant for systems in production that are
   continually evolving in small increments.

@@

\section{Analysis}
   The purpose of an information system is to make data meaningful to its users.
   Users have their own tasks and responsibilities
   and may work from different locations and on different moments.
   This collective use serves a purpose which we loosely call ``the business''.
   As a consequence, the data in a system changes continually.
   To preserve meaning, users are trying to maintain semantic constraints on the data,
   amidst of all changes that are going on around them.

   Information systems are typically used by distributed actors (both users and computers) who continually work with data.
   Every system contains a dataset, which represents the state of the system.
   Events that the system detects may cause the state to change.
   To keep our theory technology independent, we will assume that a dataset contains triples.
   This makes our theory valid for any kind of database that triples can represent,
   such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.

@@

   We will assume that the schema of a system contains constraints on its dataset.
   These constraints represent the semantics of the system~\cite{Spivak2012}.
   This allows us to be explicit about ``preserving the meaning as much as possible''.
   Our approach works with relations in a way similar to Alloy~\cite{Alloy2006} and allegories%
\footnote{Allegories are a specific type of categories, which the reader does not need to understand for the purpose of this paper}~\cite{Zielinski2013}.

   In this work we use the Ampersand compiler~\cite{Joosten-JLAMP2018},
   to generate information systems.
   A script in the Ampersand language represents the system's semantics by constraints only.
   We don't have to formalize the semantics from specifications or from source code
   because the constraints are already explicitly available in the Ampersand script.
   A system generated by the Ampersand compiler will keep these constraints satisfied at all times.
   This makes Ampersand an obvious platform for validating our theory.
   An Ampersand script contains just enough information to generate a complete system,
   which means that a classical database schema (i.e.\ data structure plus semantics) can be extracted from the Ampersand script.
   For the purpose of this paper we can equate an Ampersand script with the schema of the generated system.

   For data migration, complications arise from accumulated data pollution in the old system.
   We distinguish data pollution in the following categories:
\begin{itemize}
   \item data pollution that cannot be captured in constraints on the dataset,
   such as a street address that has become obsolete because the system never registered the change of address.
   \item data pollution that violates a constraint in the schema.
   \item data pollution that violates a constraint that is not in the schema.
   \item data pollution that is a consequence of an erroneous constraint in the schema.
\end{itemize}
   The first category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
   In some cases, constraints on data can be formulated,
   so the system can help the user and simplify (or eliminate) the corresponding working procedure.
   So, such constraints may help to move this type of pollution to one of the other categories.
   The second category is about violations of constraints from the schema.
   Such violations do not occur in our theory because the old system was generated by Ampersand,
   so that system will keep all constraints from the schema satisfied at all times.
   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
   The third category, data pollution that is not captured by any of the constraints mentioned in the schema,
   can be dealt with by adding constraints in the new system.
   In that case, a migration engineer must ensure that the new system can start with data that satisfies all constraints in its schema.
   So some human intervention may be necessary in specific cases.
   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in the new system.

\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.35]{Migration.png}
   \end{center}
\caption{Data migration}
\label{fig:event flow}
\end{figure}
   This illustrates that automating data migrations has its limits.
   We cannot eliminate the need for interventions by a migration engineer entirely.
   If the semantics encoded in the new schema differs from the old schema,
   a migration engineer must make choices.
   If the interpretation of data changes,
   a migration engineer must remove obsolete constraints from the old system and introduce new constraints in the new system.
   The consequences of such changes may require changes in data as well.
   The migration engineer may even want to adapt existing data of which the constraints in the schema have not changed.
   All of this requires a third schema, which specifies the migration itself.
   This migration schema preserves data from the old system, it may introduce new data automatically (generated from the old dataset),
   and it may require users to introduce new data (which will take time). 
   It must contain the old schema, but it also contains the new schema and possibly some new data, all of which a migration engineer must specify.
   As the migration schema must distinguish between the old dataset and the new dataset,
   the migration must be based on the disjoint union of the two.
   So the idea of a migration schema is to compute the disjoint union of the old and the new schemas
   and to present it to the migration engineer in the form of source code,
   so the migration engineer can change everything she wants to suit the particulars of the data migration.

@@

   Atoms serve as data elements.
   Atoms are values without internal structure of interest, meant to represent data elements in a database.
   From a business perspective, atoms represent concrete items of the world,
   such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
   By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
   All atoms are taken from an infinite set called $\Atoms$.

@@

   The relation $\inst:\Pair{\Atoms}{\Concepts}$ is the instance relation between atoms and concepts.
   The expression $a\ \inst\ C$ means that atom $a$ is an \emph{instance} of concept $C$.
   This relation is used in the type system, in which $\inst$ assigns a type to every atom.
   This makes $\inst$ a total relation:
\begin{eqnarray}
   \forall a\in\atoms\ \exists C\in\Concepts: a\ \inst\ C
\label{eqn:concepts}
\end{eqnarray}

@@

   Triples serve to represent data.
   A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
   A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
\begin{eqnarray}
   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
   \label{eqn:type}
\end{eqnarray}
   For example, $\triple{\text{\atom{Peter}}}{\declare{\id{phone}}{\tt Person}{\tt TelephoneNumber}}{\text{\atom{074238991}}}$ is a triple.
   Equation~\ref{eqn:type} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
   In practice, users can say that Peter has telephone number 074238991.
   So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
   This ``meaning from practice'' has no consequences in the formal world.
   We leave it entirely up to a user to attach a practical meaning to a triple.

   $\triples$ is a finite set of triples and $\inst$ is the instance relation.
   In case we need to be explicit about the dataset,
   we will write $\triples_{\dataset}$ and $\inst_{\dataset}$ to refer to $\triples$ and $\inst$ respectively.
   Every dataset is an element of an infinite set called $\Dataset$.
   To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.

@@

   In order to do static type checking,
   all concepts and relations must be ``known'' in the schema:
\begin{eqnarray}
   \declare{n}{A}{B}\in\rels&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
   \label{eqn:relationsIntroduceConcepts}\\
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}
\end{eqnarray}

   A rule in a schema serves to constrain the dataset, to ensure its semantic integrity.
   For every rule $u$ in the schema, we assume there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
   that represents the set of violations of rule $u$ in any dataset.
   We also assume a function $\Sign{u} : \Dataset\rightarrow\Pair{\Concepts}{\Concepts}$
   that represents the signature of a rule, i.e. the type of atoms in violations:
\begin{eqnarray}
   \sign{u}{\dataset}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
   \label{eqn:typed violations}
\end{eqnarray}
   To determine whether rule $u$ is satisfied in dataset $\dataset$,
   we define $\sat{u}{\Dataset}$:
\begin{eqnarray}
   \sat{u}{\dataset}&\Leftrightarrow&\viol{u}{\dataset}=\emptyset
   \label{eqn:sat}
\end{eqnarray}
   Violations are used to produce meaningful error messages and to trigger changes in the dataset to restore satisfaction of that rule.

   We assume a special kind of rule that is used for specialization.
   The rule $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
\begin{equation}
   \label{eqn:specialization}
   \sat{A\isa B}{\dataset}\ \Leftrightarrow\ \forall a: a\ \inst\ A\rightarrow a\ \inst\ B
\end{equation}
   We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
   The specialization rules in a schema form a partial order $\isa : \Pair{\concepts}{\concepts}$,
   meaning that $\isa$ is reflexive, transitive, and antisymmetric.
   Specialization rules are special for two reasons:
   Firstly, the system keeps $A\isa B$ satisfied automatically by changing $\inst$ whenever needed.
   This means that a system that contains rule $A\isa B$ can never be brought into a state in which $\id{sat}_{A\isa B}$ is False.
   This allows for certain run-time optimizations, which are beyond the scope of this paper.
   A second property that sets specialization rules apart, is that they play a role in the Ampersand compiler.
   For example, code that determines equality between atoms only works for atoms that are an instance of the same concept.
   To ensure that this does not result in problems at run-time, users are required to write all specialization rules explicitly.

@@

\subsection{Information Systems}
\label{sct:Information Systems}
   This section defines the notion of information system.
   As before, we will suffix the elements of this definition if the situation requires it.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\functionality,\roles,\maintain\ra$, in which
\begin{itemize}
   \item dataset $\dataset$ is defined as in section~\ref{sct:Datasets};
   \item schema $\schema$ is defined as in section~\ref{sct:Schemas};
   \item $\functionality$ represents the functionality of the system, which is typically software;
   \item $\roles$ is a set of roles;
   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$.
\end{itemize}
\end{definition}
   A \define{role} is a name that identifies a group of users.
   It serves as a placeholder for a person or a machine (i.e. an actor) who can change the dataset.
   The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.
   In the sequel, when we talk about a role we actually mean an actor who fulfills that role in the system.

   Every rule in a schema is meant to be kept satisfied invariably.
   The only way to break this invariance is to remove the rule from the schema.
   So the "normal" state of a system is that its dataset $\dataset$ contains no violations of any rule in the schema:
\begin{equation}
   \forall u\in\rules:\ \sat{u}{\dataset}\label{eqn:satisfaction}
\end{equation}
   However, the system continually registers events that insert or delete triples in its dataset,
   which may cause rules to be satisfied no longer.
   This is a temporary breach of the invariance of the affected rules,
   which requires a reaction to restore invariance.
   Example: if a rule says that every request must be acknowledged, the receipt of a request (which is an event) will violate this rule.
   Sending a message that acknowledges this receipt (another event) will cause the rule to be satisfied again.
   During the time between the receipt of the request and sending the acknowledgment, the rule is (temporarily) not satisfied.

   In this paper we assume two ways of enforcing satisfaction of rules: automatic or manual.
   Manual enforcement is specified by the developer by assigning a rule $u$ to a role $o$.
   Any actor with role $o$, whether person or machine,
   can restore invariance of rule $u$ by inserting or deleting the right triples to compensate for the violations.
   For this reason, we assume that every rule is assigned to at least one role by means of the relation $\maintain$.
\begin{equation}
   \forall u\in\rules\ \exists o\in\roles:\ o\ \maintain\ u\label{eqn:maintain}
\end{equation}
   $o\ \maintain\ u$ means that role $o$ must keep rule $u$ satisfied, i.e.\ to keep $\sat{u}{\dataset}$ true.
   As long as rule $u$ is not satisfied, its violations won't go away, stimulating the actor to do something about it.

   Automatic enforcement is specified by the developer with a special syntax,
   which specifies not only the rule, but also the way to restore invariance.
   The system contains an engine that restores invariance of all automatically enforced rules without unneccessary delay.
   Whenever the dataset violates an enforce rule,
   this engine will make other changes to the dataset to satisfy that rule.
   Hence, the violation of the rule exists during a small, finite amount of time.

   The system contains a dedicated role for the enforcement engine,
   so every rule has at least one role in the relation $\maintain$, means that all roles together are keeping all rules satisfied.
   (\verb-@sjcjoosten-, what if $\maintain$ relates both a human role and the enforcement engine role to the same rule?)

@@

   The example system also has a second relation that states which modules are part of which course.
   It is constrained by \verb-[UNI]- to be univalent.
   This means that every \verb-Module- is part of at most one \verb-Course-.
\begin{verbatim}
RELATION isPartOf[Module*Course] [UNI] =
[ ("Finance", "Management")
; ("Business Rules", "Business IT")
; ("Business Analytics", "Business IT")
; ("IT-Governance", "Management")
]
\end{verbatim}
   The third and last relation states which students are enrolled for which module.
   We leave it empty for now.
\begin{verbatim}
RELATION isEnrolledFor[Student*Module]
\end{verbatim}
   We also define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that a student takes.
   In Ampersand, which is a syntactically sugared form of relation algebra,
   we give each rule a name and declare a role to preserve its invariance:
\begin{verbatim}
RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
ROLE Administrator MAINTAINS EnrollRule
\end{verbatim}
   The actual constraint, \verb#isEnrolledFor |- takes;isPartOf~#, is written in logic as:
\begin{equation}
   \begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
\end{array}
\label{eqn:example isEnrolledFor}
\end{equation}
